\vspace{-1in} Demystifying the Computer Science \\PhD Admission in US Universities\\{\large A Guide for International Students

### 2023-09-19 demystify.tex
Updates from Overleaf

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those from smaller countries such as Vietnam, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted is difficult compared to CS PhD programs in other countries.

@@

# - Asking professors if they can get in or if they take students.\\
#   > Many international students will write professors asking if they can
#   get in, or if they are taking students. Unlikely anyone will reply to
#   such email because the short answer is we simply cannot evaluate your
#   profile ourselves, it needs to go to the whole committee as explained.
#   Even if a professor is seeking for students (almost everyone is
#   looking for good students), they simply cannot take one in without
#   going through the application process.

# - See this link for a large list of [[https://github.com/dynaroars/dynaroars.github.io/wiki/Funding][funding sources for PhD students]]

# ** Self-funded
# - Some PhD students are funded through their company
#   - The students work full time for the company and do their PhD part-time
#   - Typically will take longer (because you have to work full time) and require some serious commitment from the students

# - Some PhD students are funded through country, e.g., GMU has several students from Saudi Arabia that are funded through their government.

# ** Others  
# - There are also various kind of one-time thing and small awards from industry or organization, e.g., a $2000 award for the summer.
Unlike the B.A., where you or your parents pay many tens of thousands of dollars,
or the M.S., where you typically work as a teaching assistant and possibly continue
to pay many tens of thousands of dollars, the PhD is a time where funding is not a
concern to you. At most schools, you will not pay tuition during the time that you
are getting a PhD. Typically, you will also receive a living stipend – on the order
of $2000 per month, from which you will pay your living expenses. Ideally, your
only responsibility will be research. This is called doing an RAship (Research
Assistantship).
The PhD is a tremendous opportunity. You get to pick an advisor in any re-
search area you like and then you get to do research in that area, receive mentoring,
think deeply on problems, publish papers, become famous, while paying zero tu-
ition for 6 years and receiving a salary. Your advisor is paying for this opportunity
by writing grant proposals to companies and to the government to ask for fund-
ing. A single graduate student can cost an advisor upwards of 80K per year (given
the cost of tuition, the stipend, the overhead tax charged by the school, cost of
equipment and physical space, etc.).
Important note 1: At most schools, you can only do an RAship if you have
an advisor who has funding for you. Since some advisors don’t apply for grants
or are in areas which aren’t well-funded, you may have to work as a teaching
assistant every semester to get your stipend. This is called a TAship (Teaching
Assistantship). When I was a graduate student, I had a few friends who were
forced to TA 13 semesters, to fund their way through school! Alternatively, you
will have to restrict your choice of advisors to those who have funding. At CMU,
every PhD student is guaranteed a stipend plus tuition regardless of which advisor
she chooses to work with.
Important note 2: There are many companies and government organizations
which offer Graduate Fellowships for PhD students. If you are lucky enough
to get one of these, they will cover most of your way through graduate school,
and you will never have to worry about whether your advisor has funding or not.
Details about graduate fellowships will be discussed in Section 4.

* Links
:PROPERTIES:
:CUSTOM_ID: links
:END:
- https://idleprocess.wordpress.com/2009/12/07/why-go-to-graduate-school-and-how-to-get-into-the-program-of-your-dreams/
- https://mycsphd.org/_pages/application-parts.html
- https://emeryberger.com/admission-notes/
- https://www.elsevier.com/connect/9-things-you-should-consider-before-embarking-on-a-phd
- https://grad.uchicago.edu/wp-content/uploads/2019/08/Applying-to-PhD-Programs-Guide-2019.pdf
- https://www.pathwaystoscience.org/pdf/CIC_GradSchoolGuide.pdf
- https://www.science.org/content/article/applying-phd-these-10-tips-can-help-you-succeed
- https://chrisblattman.com/blog/2022/03/25/faqs-on-phd-applications/
- https://www.cs.cmu.edu/~harchol/gradschooltalk.pdf

# * Types of Funding for CS PhDs

# * How does GRA work?
# While you might get a 9-month $28K GRA stipend,  your advisor typically has to budget at least twice that much.  Because they have to pay your tuitition, insurance, and overhead that the university takes.  The chart below shows that each CS GRA costs about $70K at GMU.

# - Example budget of a PhD student in CS at GMU.  Using stipend and tuition rate in 2023. Conference Registration and Travel are estimates.
# - It is also **very likely** that GMU faculty won't have to pay student tuition starting Fall 2025, when this happens then we can take out  Tuition and total cost is 46.5K.

# | Budget                                 | Cost    |                                                                                           |
# |----------------------------------------+---------+-------------------------------------------------------------------------------------------|
# | GRA (9-month)                          | 28K     |                                                                                           |
# | GRA (summer)                           | 9K      | 3-month, 20hrs/week                                                                       |
# | Fringe Benefits (7.3%)                 | 0       | for faculty mostly (non-wage expenses such as social security, medicare)                  |
# | **Total Salary**                       | 37K     |                                                                                           |
# | Health Insurance                       | 3K      | Full year                                                                                 |
# | Tuition (In-State) for Fall and Spring | 15K     | ($680/ Credit + $150/Student Fee/ Credit)* 9 credits = $7470 ($6120 + $1350) per semester |
# | **Total**                              | 18K     | Full year tuition + insurance                                                             |
# | Total Materials & Supplies             | 0       |                                                                                           |
# | Conference Registration                | 500     |                                                                                           |
# | International Travel                   | 1.8K    |                                                                                           |
# | Domestic Travel                        | 700     |                                                                                           |
# | **Total Travel**                       | 3K      |                                                                                           |
# | Total **Direct** Cost                  | 58K     | Salary 37K + Travel 3K  + Health 3K + Tuition 18K                                         |
# | F & A (MTDC)                           | 21K     | Direct Cost - GRA Salary                                                                  |
# | Total **Indirect** Cost                | 12K     | **58.9%** of MTDC                                                                         |
# | **Total** (Direct + Indirect)          | **70K** |                                                                                           |

# - F & A = Facilities & Administrative Cost Base 
# - MTDC = Modified Total Direct Cost

# * Miscs

# - Asking professors if they can get in or if they take students.\\
#   > Many international students will write professors asking if they can
#   get in, or if they are taking students. Unlikely anyone will reply to
#   such email because the short answer is we simply cannot evaluate your
#   profile ourselves, it needs to go to the whole committee as explained.
#   Even if a professor is seeking for students (almost everyone is
#   looking for good students), they simply cannot take one in without
#   going through the application process.

# - How long for a CS PhD in US ? > 5-6 years on avg.

# - Does having an MS help? > It can help if the MS gives research
#   experience or is from a more well-known school than your
#   undergraduate. But not having an MS does not hurt either. Nowadays it
#   is normal for students with bachelor degrees apply directly to
#   PhD in CS.

>>>>

\begin{abstract}
Having been involved in PhD admissions for many years, I've
realized that many \textbf{international students}, especially those from smaller countries such as Vietnam, lack a clear understanding of
the Computer Science PhD admission process at US universities. This confusion not only
discourages students from applying but also creates the perception that
getting admitted to a CS PhD program in the US is difficult compared to other countries.

@@
